MSF_Groupwork-NBA_shotlog

### 2016-10-12 MSFGroupwork-NBAHotHand.Rmd
add analysis for shooting distance accuracy

@@

#adding a column 'DATE'
df1[,"MATCHUP"] <- as.character(df1[,"MATCHUP"])
for (i in 1:nrow(df1)) df1$DATE[i] <- as.character(as.Date(unlist(strsplit(df1$MATCHUP[i]," -"))[1], format= "%B %d, %Y"))

restdays<-cast(df1, DATE+ HOME_TEAM + AWAY_TEAM + WINNER ~ FGM)
```

@@

## Home Advantage analysis
###Introduction

This area of anlysis focuses on investigation of whether playing at 'home' gives the team advantage comparing to playing 'away'.

Literature review demonstrates that

The NULL Hypothesis is:

@@

##Defender Proximity Analysis  
###Closest Defender Distance vs Shot Accuracy
####Introduction  
This analysis wants to calculate whether there was any significant effect from closest defender distance from a shooter with the shot outcome. The hypothesis is the further a defender from the shooter, then it will be more likely to increase the chance of scoring. This is because the shooter is practically in "free" condition without any physical interference.

###Methods  
The steps to calculate this measurement are as follows:  
  1. Assign shot result to determine whether the shot is succeed (1) or missed (0)  
  2. Select distinct of **all** closest defender distance  
  3. Count the occurence fOr every distance  
  4. Calculate how many succesfull shot scored from every distance point  
  5. Calculate accuracy by dividing successfull shot scored with distance occurence for every distance point  
  6. Plot the result
```{r}
#Calculate accuracy against defender distance
#Group the defender distance level
distance_level <- as.data.frame(table(factor(ClosestDefenderDistance)))
head(distance_level)

@@

#Plotting the result
library(ggplot2)
library(scales)
ggplot(data = distance_level, aes(x = distance_level$Var1, y = distance_level$Accuracy)) +
  geom_point(aes(x = distance_level$Var1, y = distance_level$Accuracy)) +
  scale_y_continuous(labels = percent, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0), limits = c(0, 1)) +
  scale_x_discrete("Closest Defender Distance", breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24)) +
  ylab("Shooting Accuracy (%)") +
  ggtitle("Closest Defender Distance vs Shooting Accuracy")

```
###Plot Result
  The explanation from the plot above are as follows:   
  1. If the defender stands really close to the shooter (0 to 2 feet), the accuracy varied between **42%** to **50%**  
  2. If the defender distance range from 2 (two) to 4 (four) feet, the accuracy rate for this distance range from **44%** to **49%**  
  3. If the defender distance range from 4 (four) to 6 (six) feet, the accuracy rate for this distance range from **41%** to **46%**
  4. If the defender distance range from 6 (six) to 8 (eight) feet, the accuracy rate for this distance range from **37%** to **42%**
  5. If the defender distance range from 8 (eight) to 10 (ten) feet, the accuracy rate for this distance range from **37%** to **54%**
  6. For defender distance more than 10 feet up until 16 feet, the accuracy varies greater than previous distance range, starting from **33%** to **60%**.  
  7. For the distance greater than 16 feet, the data varies between **0%** (no shots scored from that distance) to **100%** (certain successfull shot from that distance)

  Next analysis will be calculated from correlation test. If the analysis included **all** of the data points for every distance, the correlation result would be as follows:
```{r}
distance_level$Var1 <- as.numeric(distance_level$Var1)
distance_level$Var1 <- distance_level$Var1 * 0.1 - 0.1

distanceZerotoAll <- c()
accuracyDist1 <- c()
for (i in 0:nrow(distance_level)){
  distanceZerotoAll <- c(distanceZerotoAll, distance_level[i, "Var1"])
  accuracyDist1 <- c(accuracyDist1, distance_level[i, "Accuracy"])
}

dfZerotoAll <- data.frame(distanceZerotoAll, accuracyDist1)
cor.test(dfZerotoFive$distanceZerotoAll, dfZerotoFive$accuracyDist1)
```
###Correlation Test Result  
  The correlation test result produced slightly high number of correlation between defender distance with shooting accuracy for all shot distance attempt in the last seasons's NBA (0.59). If the findings made only based on this number only, there would be obvious that further defender distance will **automatically** increase shooting accuracy. However, after the analysis went into smaller piece of distance, ther result will be difference. For example, for distance range between 0 (zero) to 8 feet, the correlation measurement will be as follows:

```{r}
distanceZerotoEight <- c()
accuracyDist2 <- c()
for (i in 0:80){
  distanceZerotoEight <- c(distanceZerotoEight, distance_level[i, "Var1"])
  accuracyDist2 <- c(accuracyDist2, distance_level[i, "Accuracy"])
}

dfZerotoEight <- data.frame(distanceZerotoEight, accuracyDist2)
cor.test(dfZerotoEight$distanceZerotoEight, dfZerotoEight$accuracyDist2)
```
  Therefore,

###Discussion  
  As the analysis shown, the accuracy rate of every player in the last season was somehow **affected** by the defender distance for certain range. For defender distance starting from 0 (zero) to 8 (eight) feet For the distance greater than 16 feet, the data varies between 0% (no shots scored from that distance) to 100% (certain successfull shot from that distance). It is also shown from the dataset that only small attempt of shots were made from further defender distance affecting the accuracy calculation result. This was arguably because further defender distance did not implied shooter distance from the basket is nearer to improve the accuracy.

@@

###Shooting Distance vs Shot Accuracy
####Introduction  
This analysis wants to calculate whether is there any significant effect from closest defender distance from a shooter with the shot outcome. The hypothesis is the further a defender from the shooter, then it will be more likely to increase the chance of scoring. This is because the shooter is practically in "free" condition without any physical interference.

###Methods  
The steps to calculate this measurement are as follows:  
  1. Assign shot result to determine whether the shot is succeed (1) or missed (0)  
  2. Select distinct of **all** closest defender distance  
  3. Count the occurence fOr every distance  
  4. Calculate how many succesfull shot scored from every distance point  
  5. Calculate accuracy by dividing successfull shot scored with distance occurence for every distance point  
  6. Plot the result
```{r}
#Calculate accuracy against shooting distance
#Group the shooting distance level
shooting_distance_level <- as.data.frame(table(factor(ShotDistance)))
head(shooting_distance_level)

@@

#Calculate shooting accuracy for each distance
shooting_distance_level$Accuracy <- round(shooting_distance_level$SCORED / shooting_distance_level$Freq, 2)

#Plotting the result
library(ggplot2)
ggplot(data = shooting_distance_level, aes(x = shooting_distance_level$Var1, y = shooting_distance_level$Accuracy)) +
  geom_point(aes(x = shooting_distance_level$Var1, y = shooting_distance_level$Accuracy))

#Data Preparation
#Change the value of shot result column to numeric
df[, "SHOT_RESULT"] <- as.character(df[, "SHOT_RESULT"])
for (j in 1:nrow(df)){
  df[j, "SHOT_RESULT"] <- ifelse (df[j, "SHOT_RESULT"] == "made", 1, 0)
}

#set the value to numeric or integer accordingly
df[, "SHOT_RESULT"] <- as.numeric(df[, "SHOT_RESULT"])
df[, "PTS_TYPE"] <- as.integer(df[, "PTS_TYPE"])
df[, "CLOSE_DEF_DIS"] <- as.numeric(df[, "CLOSE_DEF_DIST"])

#select data with only scored result
#Point type (2 or 3 points) and distance
point_type <- c()
distance <- c()
for (z in 1:nrow(df)){
  if(df[z, "SHOT_RESULT"] == 1 & df[z, "PTS_TYPE"] >= 1){
    point_type <- c(point_type, df[z, "PTS_TYPE"])
    distance <- c(distance, df[z, "CLOSE_DEF_DIST"])
  }
}

#combine point type and distance
valid_distance = data.frame(point_type, distance)

#calculate correlation between defender distance and points made (2 or 3 points)
cor(point_type, distance)

#plot the diagram for this correlation
library(ggplot2)
ggplot(data = valid_distance, aes(x = point_type, y = distance, group = point_type)) + 
  geom_boxplot() + stat_summary(fun.y = mean, geom = "point", shape = 5, size = 1) + 
  ggtitle("Points Scored based on Distance from Defender")

#calculate correlation between defender distance and shot result
cor(df[, "SHOT_RESULT"], df[, "CLOSE_DEF_DIS"])

#--------------------------------------------------------------------------------#

#Further analysis
#select distance data from 2 points shot scored and the players who scored it
distance2 <- c()
players2 <- c()
for (z in 1:nrow(df)){
  if(df[z, "SHOT_RESULT"] == 1 & df[z, "PTS_TYPE"] == 2){
    distance2 <- c(distance2, df[z, "CLOSE_DEF_DIST"])
    players2 <- c(players2, as.character(df[z, "player_name"]))
  }
}

#combine 2 point players and defender distance
player2point <- data.frame(players2, distance2)
summary(player2point)

#count players who scored 2 points
counter <- as.data.frame(table(players2))

#count 5 top players
top5Players2 <- counter[rev(order(counter$Freq)), "players2"][1:5]
top5Counter <- counter[rev(order(counter$Freq)), "Freq"][1:5]

top5_2points <- data.frame(top5players2, top5Counter)

#plot the graph
library(ggplot2)
ggplot(data = top5_2points, aes(x = top5Players2)) +
  geom_bar(aes(x = top5players2, y = top5Counter), stat = "identity")

#----------------------------------------------------------------#

#select distance data from 3 points shot scored and the players who scored it
distance3 <- c()
players3 <- c()
for (z in 1:nrow(df)){
  if(df[z, "SHOT_RESULT"] == 1 & df[z, "PTS_TYPE"] == 3){
    distance3 <- c(distance3, df[z, "CLOSE_DEF_DIST"])
    players3 <- c(players3, as.character(df[z, "player_name"]))
  }
}

#combine 3 point players and defender distance
player3point <- data.frame(players3, distance3)
summary(player3point)

#count players who scored 3 points
counter3 <- as.data.frame(table(players3))

#count 5 top players
top5Players3 <- counter3[rev(order(counter3$Freq)), "players3"][1:5]
top5Counter3 <- counter3[rev(order(counter3$Freq)), "Freq"][1:5]

top5_3points <- data.frame(top5Players3, top5Counter3)

#plot the graph
library(ggplot2)
ggplot(data = top5_3points, aes(x = top5Players3)) +
  geom_bar(aes(x = top5Players3, y = top5Counter3), stat = "identity")

@@

```{r}
#Summary of total points
ptssum<- cast(df1, GAME_ID1+ player_id + player_name  ~ PTS); head(ptssum)
#creating "total points per player, per game" column
ptssum$totalpts<- ptssum[,"2"]*2 + ptssum[,"3"]*3

@@

#Compare barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar() + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar() + labs(title = 'Klay Thompson')
grid.arrange(g1, g2, ncol = 2)
#Klay Thompson is an exception, had 13 streaks in a row! We cannot find a similar distribution in our sample, so we can assume that he is likely to have a hot hand. Fearther statistical analysis (logistic regression) should be done in order to reach a conclusion.
```

>>>>

### Home Advantage analysis

@@

@@

@@

##Defender Proximity Analysis  
###Closest Defender Distance vs Shot Accuracy
####Introduction  
This analysis would calculate whether there was any significant effect from closest defender distance from a shooter with the shot outcome. The hypothesis is the further a defender from the shooter, then it will be more likely to increase the chance of scoring. This is because the shooter is practically in "free" condition without any physical interference.

###Methods  
The steps to calculate this measurement are as follows:  
  1. Assign shot result to determine whether the shot is succeed (1) or missed (0)  
  2. Select distinct of **all** closest defender distance  
  3. Count the occurence fOr every distance  
  4. Calculate how many succesfull shot scored from every distance point  
  5. Calculate accuracy by dividing successfull shot scored with distance occurence for every defender distance point  
  6. Plot the result
```{r}
#Calculate accuracy against defender distance
#Group the defender distance level
distance_level <- as.data.frame(table(factor(ClosestDefenderDistance)))
head(distance_level)

@@

#Plotting the result
library(ggplot2)
library(scales)
ggplot(data = distance_level, aes(x = distance_level$Var1, y = distance_level$Accuracy)) +
  geom_point(aes(x = distance_level$Var1, y = distance_level$Accuracy)) +
  scale_y_continuous(labels = percent, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0), limits = c(0, 1)) +
  scale_x_discrete("Closest Defender Distance", breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24)) +
  geom_smooth() +
  ylab("Shooting Accuracy (%)") +
  ggtitle("Closest Defender Distance vs Shooting Accuracy")

```
###Plot Result
  The explanation from the plot above are as follows:   
  1. If the defender stands really close to the shooter (0 to 2 feet), the accuracy varied between **42%** to **50%**  
  2. If the defender distance range from 2 to 4 feet, the accuracy rate for this distance range from **44%** to **49%**  
  3. If the defender distance range from 4 to 6 feet, the accuracy rate for this distance range from **41%** to **46%**
  4. If the defender distance range from 6 to 8 feet, the accuracy rate for this distance range from **37%** to **42%**
  5. If the defender distance range from 8 to 10 feet, the accuracy rate for this distance range from **37%** to **54%**
  6. For defender distance more than 10 feet up until 16 feet, the accuracy varies greater than previous distance range, starting from **33%** to **60%**.  
  7. For the distance greater than 16 feet, the data varies between **0%** (no shots scored from that distance) to **100%** (certain successfull shot from that distance)

  Next analysis will be calculated from correlation test. If the analysis included **all** of the data points for every distance, the correlation result would be as follows:
```{r}
#change distance level to numeric type for correlation test function
distance_level$Var1 <- as.numeric(distance_level$Var1)
distance_level$Var1 <- distance_level$Var1 * 0.1 - 0.1 #to keep the distance point as it is

#populate data points from data frame
distanceZerotoAll <- c()
accuracyDist1 <- c()
for (i in 0:nrow(distance_level)){
  distanceZerotoAll <- c(distanceZerotoAll, distance_level[i, "Var1"])
  accuracyDist1 <- c(accuracyDist1, distance_level[i, "Accuracy"])
}

#combine data frame
dfZerotoAll <- data.frame(distanceZerotoAll, accuracyDist1)

#calculate correlation
cor.test(dfZerotoFive$distanceZerotoAll, dfZerotoFive$accuracyDist1)
```
###Correlation Test Result  
  The correlation test result produced slightly high number of correlation between defender distance with shooting accuracy for all shot distance attempt in the last seasons's NBA (0.59). If the findings made only based on this number only, there would be obvious that further defender distance will **automatically** increase shooting accuracy. However, after the analysis went into smaller piece of distance, ther result will be difference. For example, for distance range between 0 (zero) to 8 feet, the correlation measurement will be as follows:

```{r}
#populate data from data frame
distanceZerotoEight <- c()
accuracyDist2 <- c()
for (i in 0:80){
  distanceZerotoEight <- c(distanceZerotoEight, distance_level[i, "Var1"])
  accuracyDist2 <- c(accuracyDist2, distance_level[i, "Accuracy"])
}

#combine data frame
dfZerotoEight <- data.frame(distanceZerotoEight, accuracyDist2)

#calculate correlation
cor.test(dfZerotoEight$distanceZerotoEight, dfZerotoEight$accuracyDist2)
```
  As the analysis shown, the correlation measurement between 0 (zero) to 8 feet would be negative (-0.40). This result was different with the previous calculation done for all data points (0.59).

###Discussion  
  As the analysis shown, the accuracy rate of every player in the last season was somehow **affected** by the defender distance for certain range. For defender distance starting from 0 (zero) to 8 (eight) feet For the distance greater than 16 feet, the data varies between 0% (no shots scored from that distance) to 100% (certain successfull shot from that distance). It is also shown from the dataset that only small attempt of shots were made from further defender distance affecting the accuracy calculation result. However, from the correlation perspective, the result would be different if the analysis included all of the data points versus only included short distance data points. This was arguably because further defender distance did not implied shooter distance from the basket is nearer to improve the accuracy and also there were not many players to perform shooting attempt automatically although their opponent was standing far from the shooter

@@

###Shooting Distance vs Shot Accuracy
####Introduction  
This analysis would calculate whether was there any significant effect from shooting distance from basket with the shot outcome. The hypothesis is the further a shooter from the basket, then the success rate would be decreased. This was because when the shooter was practically in closer shooting range, the shooter shooting's result would be more accurate.

###Methods  
The steps to calculate this measurement are as follows:  
  1. Assign shot result to determine whether the shot is succeed (1) or missed (0)  
  2. Select distinct of **all** shooting distance  
  3. Count the occurence fOr every distance  
  4. Calculate how many succesfull shot scored from every distance point  
  5. Calculate accuracy by dividing successfull shot scored with particular distance occurence for every shooting distance point  
  6. Plot the result
```{r}
#Calculate accuracy against shooting distance
#Group the shooting distance level
shooting_distance_level <- as.data.frame(table(factor(ShotDistance)))
head(shooting_distance_level)

@@

#Calculate shooting accuracy for each distance
shooting_distance_level$Accuracy <- round(shooting_distance_level$SCORED / shooting_distance_level$Freq, 4)

#Plotting the result
library(ggplot2)
library(scales)
ggplot(data = shooting_distance_level, aes(x = shooting_distance_level$Var1, y = shooting_distance_level$Accuracy)) +
  geom_point(aes(x = shooting_distance_level$Var1, y = shooting_distance_level$Accuracy)) +
  scale_y_continuous(labels = percent, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0), limits = c(0, 1)) +
  scale_x_discrete("Shooting Distance", breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28, 30, 32, 34, 36, 38, 40, 42, 44, 46, 48)) +
  geom_smooth() +
  ylab("Shooting Accuracy (%)") +
  ggtitle("Shooting Distance vs Shooting Accuracy")

@@

###Plot Result
  The explanation from the plot above are as follows:   
  1.  If the shooter stands really close to the basket (0 to 2 feet), the accuracy varied between **62%** to **68%**  
  2.  If the shooting distance range from 2 to 4 feet, the accuracy rate for this distance range from **61%** to **67%**  
  3.  If the shooting distance range from 4 to 6 feet, the accuracy rate for this distance range from **44%** to **58%**  
  4.  If the shooting distance range from 6 to 8 feet, the accuracy rate for this distance range from **38%** to **51%**  
  5.  If the shooting distance range from 8 to 10 feet, the accuracy rate for this distance range from **35%** to **42%**  
  6.  If the shooting distance range from 10 to 12 feet, the accuracy rate for this distance range from **34%** to **42%**  
  7.  If the shooting distance range from 12 to 14 feet, the accuracy rate for this distance range from **37%** to **47%**  
  8.  If the shooting distance range from 14 to 16 feet, the accuracy rate for this distance range from **33%** to **51%**  
  9.  If the shooting distance range from 16 to 18 feet, the accuracy rate for this distance range from **37%** to **48%**  
  10. If the shooting distance range from 18 to 24 feet, the accuracy rate for this distance range from **33%** to **43%**  
  11. If the shooting distance range from 24 to 28 feet, the accuracy rate for this distance range from **18%** to **38%**  
  12. For shooting distance more than 28 feet, the accuracy varies greater than previous distance range, starting from **0%** (no shots scored from that distance) to **50%**

  Next analysis will be calculated from correlation test. If the analysis included **all** of the data points for every distance, the correlation result would be as follows:  
```{r}
#change distance level to numeric type for correlation test function
shooting_distance_level$Var1 <- as.numeric(shooting_distance_level$Var1)
shooting_distance_level$Var1 <- shooting_distance_level$Var1 * 0.1 - 0.1 #to maintain distance data point as it is

#populate data for every shooting distance
shootingDistanceZerotoAll <- c()
shootingAccuracyDist1 <- c()
for (i in 0:nrow(shooting_distance_level)){
  shootingDistanceZerotoAll <- c(shootingDistanceZerotoAll, shooting_distance_level[i, "Var1"])
  shootingAccuracyDist1 <- c(shootingAccuracyDist1, shooting_distance_level[i, "Accuracy"])
}

dfShootingZerotoAll <- data.frame(shootingDistanceZerotoAll, shootingAccuracyDist1)
cor.test(dfShootingZerotoAll$shootingDistanceZerotoAll, dfShootingZerotoAll$shootingAccuracyDist1)
```
###Correlation Test Result  
  The correlation test result shown strong negative correlation between shooting distance and shooting accuracy (-0.87). This meant that for all data points, the further a shooter from the basket, it was unlikely the shooter's attempt was success, regardless who was the shooter. However, if the correlation test taken for certain range distance based on the diagram, for example range from 10 feet to 20 feet, the correlation result would be different as follows:

```{r}
#populate data for shooting distance in range 10 to 24 feet
shootingDistance10to20 <- c()
shootingAccuracyDist2 <- c()
for (i in 100:200){
  shootingDistance10to20 <- c(shootingDistance10to20, shooting_distance_level[i, "Var1"])
  shootingAccuracyDist2 <- c(shootingAccuracyDist2, shooting_distance_level[i, "Accuracy"])
}

dfShooting10to20 <- data.frame(shootingDistance10to20, shootingAccuracyDist2)
cor.test(dfShooting10to20$shootingDistance10to20, dfShooting10to20$shootingAccuracyDist2)
```
  The second correlation test produced a small positive correlation between range 10 to 20 feet (0.11). This result was different with previous measurement (-0.87)

###Discussion  
  Based on the diagram and correlation, it can be concluded that although the shooting distance would directly affect shooting accuracy, but the relationship between these variables are not perfectly linear. This was shown in the diagram where there was a relatively **similar** shooting accuracy for shots taken from 10 feet until 20 feet, regardless the shooter's performance. And for the success rate in longer distances, it could be arguably inferred as the shooters tried to attempt in closing game time (buzzer beater) when the shot clock or game clock nearly reach zero.

###Limitations  
 Since the dataset mostly return 0% for shooting distance greater than 30 feet, it would better if the dataset could be enriched with previous season statistics, or even the last 10 seasons to exercise in more detail about the pattern for long distance shots attempt.  
```{r}
#Summary of total points
ptssum<- cast(df1, GAME_ID1+ player_id + player_name  ~ PTS); head(ptssum)
#creating "total points per player, per game" column
ptssum$totalpts<- ptssum[,"2"]*2 + ptssum[,"3"]*3

@@

#Compare barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar() + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar() + labs(title = 'Klay Thompson')
grid.arrange(g1, g2, ncol = 2)
#Klay Thompson is an exception, had 13 streaks in a row! We cannot find a similar distribution in our sample, so we can assume that he is likely to have a hot hand.

best_players_df
